---
title: "Network plots"
author: "Mengbo Li"
date: "`r Sys.Date()`"
output: 
   html_document:
      code_folding: show
      toc: true
      toc_float:
          collapsed: false
          smooth_scroll: false
      toc_depth: 3
      number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r global options, include = FALSE, cache = FALSE}
options(max.print = "80", digits = 4)
knitr::opts_chunk$set(echo = FALSE,
                      cache = FALSE,
                      prompt = FALSE,
                      tidy = TRUE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE, 
                      fig.align = "center")
```



# Summary

Here I try to generate some network plots for the last figure. After several discussions, we have decided to create the protein network from observed data but add on the metabolomic markers in from the DE list post-hoc (send to John). 



```{r}
library(here)
library(tidyverse)
library(RColorBrewer)
library(ggsci)
library(igraph)
library(mstknnclust)
library(limma)
```


# Load data and DE lists

```{r}
# all protein data and meta data, donor topTable LV vs RV
load("data/network/211115_protData_meta_and_donorLVRVde.rda")
# all metabolomic data meta data, donor topTable LV vs RV
load("data/network/211115_metabData_meta_and_donorLVRVde.rda")
```


## Donor data

```{r}
# take out donors first
donorID <- pmeta$locBank[pmeta$condition == "Donor"]

# take top 200 DE proteins
dep <- paste("G", donor$GeneID[1:200], sep = "")
datDonor <- pt[dep, donorID]
rownames(datDonor) <- donor$Symbol[1:200]

# rescale to median and adjust by standard deviations
datDonor <- t(scale(t(datDonor)))

dim(datDonor)
```


Construct the network layout. 

```{r}
make.knn.graph <- function(D, degree.cutoff = 0.5){
  # calculate Euclidean distances between cells
  # dist <- as.matrix(dist(D))
  # or, use the correlation matrix to construct the KNN network
  dist <- 1/abs(cor(t(D), use = "pairwise.complete.obs"))
  graph <- mst.knn(dist)
  graph <- graph$network
  V(graph)$label <- V(graph)$name <- rownames(D)[as.numeric(V(graph)$name)]
  # degree of graph
  graphdegree <- igraph::degree(graph, v = V(graph), loops = TRUE, normalized = TRUE)
  # filter nodes without edge
  graph <- delete.vertices(graph, graphdegree == 0)
  # filter by degree of graph
  graph <- delete.vertices(graph, graphdegree < quantile(graphdegree, prob = degree.cutoff))
  # vertex size will be by degree
  graphdegree <- graphdegree[V(graph)$name]
  # edge weights by absolute values of the correlation
  edgeList <- get.edgelist(graph)
  edge.wt <- sapply(1:nrow(edgeList), function(i) abs(cor(D[edgeList[i, 1], ], D[edgeList[i, 2], ], use = "pairwise.complete.obs")))
  # colour edges by correlation (red for positive and blue for negative)
  corSign <- sapply(1:nrow(edgeList), function(i) sign(cor(D[edgeList[i, 1], ], D[edgeList[i, 2], ], use = "pairwise.complete.obs")))
  edge.col <- c("#EF3B2C", "#4292C6")[ifelse(corSign == -1, 2, corSign)] # red for +, blue for -
  # make a layout for visualizing in 2D
  set.seed(1)
  g.layout <- layout_with_fr(graph)
  return(list(graph = graph, 
              layout = g.layout, 
              graphdegree = graphdegree, 
              edge.col = edge.col, 
              edge.wt = edge.wt))        
}

donorknn <- make.knn.graph(datDonor, degree.cutoff = 0)
donorWalktrap <- walktrap.community(donorknn$graph, steps = 10)
table(donorWalktrap$membership)
community.list <- lapply(1:max(donorWalktrap$membership), function(i) donorWalktrap$names[donorWalktrap$membership == i])
```


```{r}
## firstly, the colour palette
ind <- c(2, 7)
color.plate <- c(brewer.pal(9, "Purples")[c(4, 6)], brewer.pal(9, "Blues")[c(5, 7)],
                 brewer.pal(11, "BrBG")[c(8, 9)], pal_material("green")(10)[ind], 
                 pal_material("lime")(10)[c(4, 8)], pal_material("orange")(10)[c(4, 9)], 
                 pal_material("yellow")(10)[ind], pal_material("red")(10)[c(2, 3)], 
                 pal_material("red")(10)[c(5, 6)], brewer.pal(11, "BrBG")[c(4, 3)], 
                 brewer.pal(11, "PiYG")[c(4, 3)])
length(color.plate)
scales::show_col(color.plate)
set.seed(1)
sub_color.plate <- color.plate[sample(1:22)]
scales::show_col(sub_color.plate)

# community highlighting colours, more transparency
# community.col <- GISTools::add.alpha(str_sub(sub_color.plate, end = 7), alpha = 0.28)
colorRampAlpha <- function(..., n, alpha) {
   colors <- colorRampPalette(...)(n)
   paste(colors, sprintf("%x", ceiling(255*alpha)), sep="")
}
community.col <- colorRampAlpha(str_sub(sub_color.plate, end = 7), 
               n = length(str_sub(sub_color.plate, end = 7)), 
               alpha = 0.2)

# colour nodes by community
vert.col <- sub_color.plate[donorWalktrap$membership]
# vert.frame.col <- vert.col
vert.frame.col <- data.frame(Symbol = names(V(donorknn$graph))) |> 
  left_join(select(donor, Symbol, logFC), by = "Symbol") |> 
  mutate(direction = ifelse(logFC > 0, "salmon", "lightblue"))
# adjust vertex sizes by nodal degree
vert.size <- donorknn$graphdegree*288
vert.size <- ifelse(vert.size < quantile(vert.size, 0.9), 4, 8)
```


Plot: 

```{r}
plot.igraph(donorknn$graph, layout = donorknn$layout, 
            # vertex
            vertex.label = V(donorknn$graph)$label, vertex.label.family = "Helvetica", 
            vertex.label.font = 2, vertex.label.cex = 0.4, vertex.shape = donorknn$ver.shape, 
            vertex.label.color = "black", 
            # vertex.color = vert.col,
            vertex.color = vert.frame.col$direction,
            vertex.size = vert.size, vertex.frame.color = vert.frame.col, 
            # edge
            edge.color = donorknn$edge.col, #"#404040", # 
            edge.width = (donorknn$edge.wt*1.5)^2, 
            # community
            mark.groups = community.list, mark.border = "#ffffff00",  mark.col = community.col)
legend("topleft", 
       legend = 1:22, # community.anno
       col = sub_color.plate, 
       pch = 16, cex = 0.6, 
       bty = "n", pt.cex = 1, 
       y.intersp = 1.5)
```




Gene set test in each cluster:

```{r, eval = FALSE}
# read in gene sets
## tft
tft <- readLines(here("data/211122_MSigDB/c3.tft.v7.4.entrez.gmt.txt"))
tft <- lapply(1:length(tft), function(i) limma::strsplit2(tft[i], "\t")[-2])
names(tft) <- sapply(tft, "[", 1)
tft <- lapply(tft, "[", -1)
length(tft)
## GO-BP
bp <- readLines(here("data/211122_MSigDB/c5.go.bp.v7.4.entrez.gmt.txt"))
bp <- lapply(1:length(bp), function(i) limma::strsplit2(bp[i], "\t")[-2])
names(bp) <- sapply(bp, "[", 1)
bp <- lapply(bp, "[", -1)
length(bp)
## GO-MF
mf <- readLines(here("data/211122_MSigDB/c5.go.mf.v7.4.entrez.gmt.txt"))
mf <- lapply(1:length(mf), function(i) limma::strsplit2(mf[i], "\t")[-2])
names(mf) <- sapply(mf, "[", 1)
mf <- lapply(mf, "[", -1)
length(mf)
```


```{r, eval = FALSE}
# genes in each cluster
geneClusters <- lapply(1:22, function(x) donorWalktrap[[x]])
geneCusters_id <- lapply(1:22, function(x) {
  filter(donor, Symbol %in% donorWalktrap[[x]])$GeneID
})

# Test a cluster on each gene set of interest by camera
testAcluster <- function(genesInCluster, geneSets) {
  index <- lapply(1:length(geneSets), function(i) {
    thisGS <- geneSets[[i]]
    ind <- intersect(paste("G", thisGS, sep = ""), 
                     paste("G", donor$GeneID, sep = ""))
    if (length(ind) < 5) res <- NA
      else {
        ind2 <- paste("G", intersect(geneSets[[i]], genesInCluster), sep = "")
        if (length(ind2) < 2) 
          res <- NA
        else {
          allStats <- dplyr::filter(donor, GeneID %in% thisGS)
          indStats <- allStats$t
          names(indStats) <- paste("G", allStats$GeneID, sep = "")
          index4test <- which(names(indStats) %in% ind2)
          res <- cameraPR(indStats, index = index4test)
        }
      }
    return(res)
  })
}

# test all clusters
testCollection <- function(clusterID = 1:18, geneCusters_id, collection = tft) {
  lapply(clusterID, function(i) {
    tmpRes <- testAcluster(geneCusters_id[[i]], collection)
    names(tmpRes) <- names(collection)
    keep <- sapply(tmpRes, length) > 1
    res <- do.call(rbind, tmpRes[keep])
    return(res)
  })
}
```


For each collection, take the top 5 annotations for each cluster: 

```{r, eval = FALSE}
donor_tft <- testCollection(1:22, geneCusters_id, tft)
donor_bp <- testCollection(1:22, geneCusters_id, bp)
donor_mf <- testCollection(1:22, geneCusters_id, mf)

donorAnno <- lapply(1:22, function(i) {
  if (is.null(donor_tft[[i]])) tftRes <- NA else {
  tfts <- arrange(donor_tft[[i]], PValue)
  ind_tfts <- tfts$PValue <= 0.05
  tftRes <- paste("TFT: ", paste(rownames(tfts)[1:min(5,sum(ind_tfts))],
                                        collapse = "; "), 
                         sep = "") %>% str_remove_all("_TARGET_GENES")}
  
  if (is.null(donor_bp[[i]])) bpRes <- NA else {
  bps <- arrange(donor_bp[[i]], PValue)
  ind_bps <- bps$PValue <= 0.05
  bpRes <- paste("GOBP: ", tolower(paste(rownames(bps)[1:min(5,sum(ind_bps))],
                                          collapse = "; ") %>% str_remove_all("GOBP_")), sep = "")}
  
  if (is.null(donor_mf[[i]])) mfRes <- NA else {
  mfs <- arrange(donor_mf[[i]], PValue)
  ind_mfs <- mfs$PValue <= 0.05
  mfRes <- paste("GOMF: ", tolower(paste(rownames(mfs)[1:min(5,sum(ind_mfs))],
                                          collapse = "; ") %>% str_remove_all("GOMF_")), sep = "")}
  
  data.frame(cluster = NA, 
             tft = tftRes, 
             gobp = bpRes, 
             gomf = mfRes)
}) %>% do.call(rbind, .)
donorAnno$cluster <- 1:max(donorWalktrap$membership)

write.csv(donorAnno, file = here("data/211122_MSigDB/211122_donor_network_cluster_annotations.csv"),
          row.names = FALSE)
```


## ICM data

```{r}
# Load ICM topTables
icmProt <- read.csv(here("data/network/211105_Proteomics_ICM_LVvsRV_DE.csv"))
icmID <- pmeta$locBank[pmeta$condition == "ICM"]

# take top 200 proteins
dep <- paste("G", icmProt$GeneID[1:200], sep = "")
datIcm <- pt[dep, icmID]
rownames(datIcm) <- icmProt$Symbol[1:200]

# rescale to median and adjust by standard deviations
datIcm <- t(scale(t(datIcm)))
dim(datIcm)
```


```{r}
icmknn <- make.knn.graph(datIcm, degree.cutoff = 0)
icmWalktrap <- walktrap.community(icmknn$graph, steps = 10)
table(icmWalktrap$membership)
community.list <- lapply(1:max(icmWalktrap$membership), function(i) icmWalktrap$names[icmWalktrap$membership == i])
```



```{r}
set.seed(4)
sub_color.plate <- color.plate[sample(1:22)][1:18]
scales::show_col(sub_color.plate)

# community highlighting colours, more transparency
community.col <- GISTools::add.alpha(str_sub(sub_color.plate, end = 7), alpha = 0.28)

# colour nodes by community
vert.col <- sub_color.plate[icmWalktrap$membership]

vert.frame.col <- vert.col

vert.frame.col <- data.frame(Symbol = names(V(icmknn$graph))) |> 
  left_join(select(icmProt, Symbol, logFC), by = "Symbol") |> 
  mutate(direction = ifelse(logFC > 0, "salmon", "lightblue"))

# adjust vertex sizes by nodal degree
vert.size <- icmknn$graphdegree*288
vert.size <- ifelse(vert.size < quantile(vert.size, 0.9), 4, 8)
```

Plot: 

```{r}
plot.igraph(icmknn$graph, layout = icmknn$layout, 
            # vertex
            vertex.label = V(icmknn$graph)$label, vertex.label.family = "Helvetica", 
            vertex.label.font = 2, vertex.label.cex = 0.4, vertex.shape = icmknn$ver.shape, 
            vertex.label.color = "black", 
            # vertex.color = vert.col, 
            vertex.color = vert.frame.col$direction, 
            vertex.size = vert.size, vertex.frame.color = vert.col, 
            # edge
            edge.color = "#404040", edge.width = (icmknn$edge.wt*1.5)^2, 
            # community
            mark.groups = community.list, mark.border = "#ffffff00")
            # ,  mark.col = community.col)
legend("topleft", 
       legend = 1:18, # community.anno
       col = sub_color.plate, 
       pch = 16, cex = 0.6, 
       bty = "n", pt.cex = 1, 
       y.intersp = 1.5)
```


For each collection, take the top 5 annotations for each cluster: 

```{r}
# genes in each cluster
geneClusters <- lapply(1:18, function(x) icmWalktrap[[x]])
geneCusters_id <- lapply(1:18, function(x) {
  filter(donor, Symbol %in% icmWalktrap[[x]])$GeneID
})

icm_tft <- testCollection(1:18, geneCusters_id, tft)
icm_bp <- testCollection(1:18, geneCusters_id, bp)
icm_mf <- testCollection(1:18, geneCusters_id, mf)

icmAnno <- lapply(1:18, function(i) {
  if (is.null(icm_tft[[i]])) tftRes <- NA else {
  tfts <- arrange(icm_tft[[i]], PValue)
  ind_tfts <- tfts$PValue <= 0.05
  tftRes <- paste("TFT: ", paste(rownames(tfts)[1:min(5,sum(ind_tfts))],
                                        collapse = "; "), 
                         sep = "") %>% str_remove_all("_TARGET_GENES")}
  
  if (is.null(icm_bp[[i]])) bpRes <- NA else {
  bps <- arrange(icm_bp[[i]], PValue)
  ind_bps <- bps$PValue <= 0.05
  bpRes <- paste("GOBP: ", tolower(paste(rownames(bps)[1:min(5,sum(ind_bps))],
                                          collapse = "; ") %>% str_remove_all("GOBP_")), sep = "")}
  
  if (is.null(icm_mf[[i]])) mfRes <- NA else {
  mfs <- arrange(icm_mf[[i]], PValue)
  ind_mfs <- mfs$PValue <= 0.05
  mfRes <- paste("GOMF: ", tolower(paste(rownames(mfs)[1:min(5,sum(ind_mfs))],
                                          collapse = "; ") %>% str_remove_all("GOMF_")), sep = "")}
  
  data.frame(cluster = NA, 
             tft = tftRes, 
             gobp = bpRes, 
             gomf = mfRes)
}) %>% do.call(rbind, .)
icmAnno$cluster <- 1:18

write.csv(icmAnno, file = here("data/211122_MSigDB/211122_ICM_network_cluster_annotations.csv"), 
          row.names = FALSE)
```










## DCM data

```{r}
# Load DCM topTables
dcmProt <- read.csv(here("data/network/211105_Proteomics_DCM_LVvsRV_DE.csv"))
dcmID <- pmeta$locBank[pmeta$condition == "DCM"]

# take top 100 proteins 
dep <- paste("G", dcmProt$GeneID[dcmProt$adj.P.Val <= 0.05], sep = "")
datDcm <- pt[dep, dcmID]
rownames(datDcm) <- dcmProt$Symbol[dcmProt$adj.P.Val <= 0.05]

# rescale to median and adjust by standard deviations
datDcm <- t(scale(t(datDcm)))

dim(datDcm)
```


```{r}
dcmknn <- make.knn.graph(datDcm, degree.cutoff = 0)
dcmWalktrap <- walktrap.community(dcmknn$graph, steps = 10)
table(dcmWalktrap$membership)
community.list <- lapply(1:max(dcmWalktrap$membership), function(i) dcmWalktrap$names[dcmWalktrap$membership == i])
```

Colour codes: 

```{r}
sub_color.plate <- color.plate[seq(1, 7, 2)]

# community highlighting colours, more transparency
community.col <- GISTools::add.alpha(str_sub(sub_color.plate, end = 7), alpha = 0.28)

# colour nodes by community
vert.col <- sub_color.plate[dcmWalktrap$membership]
vert.frame.col <- vert.col

vert.frame.col <- data.frame(Symbol = names(V(dcmknn$graph))) |> 
  left_join(select(dcmProt, Symbol, logFC), by = "Symbol") |> 
  mutate(direction = ifelse(logFC > 0, "salmon", "lightblue"))

# adjust vertex sizes by nodal degree
vert.size <- dcmknn$graphdegree*8
```



Plot: 

```{r}
plot.igraph(dcmknn$graph, layout = dcmknn$layout, 
            # vertex
            vertex.label = V(dcmknn$graph)$label, vertex.label.family = "Helvetica", 
            vertex.label.font = 2, vertex.label.cex = 0.4, vertex.shape = dcmknn$ver.shape, 
            vertex.label.color = "black", 
            # vertex.color = vert.col[2], 
            vertex.color = vert.frame.col$direction, 
            vertex.size = vert.size, vertex.frame.color = vert.col[2], 
            # edge
            edge.color = "#404040", edge.width = (dcmknn$edge.wt*1.8)^2)
```









<br><br><br><br>
# Session information

```{r}
sessioninfo::session_info()
```

